---
title: >
  ggcoverage User Guide
author: 
- name: Yabing Song
  affiliation: 
  - &id1 School of Life Sciences, Tsinghua University
  email: songyb18@mails.tsinghua.edu.cn
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('ggcoverage')`"
abstract: >
  The goal of `ggcoverage` is to simplify the process of visualizing genome coverage. It contains functions to load data from BAM, BigWig or BedGraph files, create genome coverage plot, add various annotation to the coverage plot, including base and amino acid annotaion, GC annotation, gene annotation, transcript annotation, ideogram annotation and peak annotation.
output: 
  BiocStyle::html_document:
    toc_depth: 4
    toc_float: true
fig_caption: TRUE
vignette: >
  %\VignetteIndexEntry{ggcoverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE, warning=FALSE}
library(knitr)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())

# set dpi
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=60
)
```

# Getting started

`ggcoverage` is an R package distributed as part of the [CRAN](https://cran.r-project.org/).
To install the package, start R and enter:
  
```{r install, eval=FALSE}
# install via CRAN
install.package("ggcoverage")

# install via Github
# install.package("remotes")   #In case you have not installed it.
remotes::install_github("showteeth/ggcoverage")
```

Once `ggcoverage` is installed, it can be loaded by the following command.

```{r library, message=FALSE, warning=FALSE}
library("rtracklayer")
library("ggcoverage")
```

# Introduction
<img src = "../man/figures/ggcoverage.png" align = "right" width = "200"/>

  The goal of `ggcoverage` is simplify the process of visualizing genome coverage. It contains three main parts:

* **Load the data**: `ggcoverage` can load BAM, BigWig (.bw), BedGraph files from various NGS data, including WGS, RNA-seq, ChIP-seq, ATAC-seq, et al.
* **Create genome coverage plot**
* **Add annotations**: `ggcoverage` supports six different annotations:
  * **Base and amino acid annotaion**: Visualize genome coverage at single-nucleotide level with bases and amino acids.
  * **GC annotation**: Visualize genome coverage with GC content
  * **gene annotation**: Visualize genome coverage across whole gene
  * **transcription annotation**: Visualize genome coverage across different transcripts
  * **ideogram annotation**: Visualize the region showing on whole chromosome
  * **peak annotation**: Visualize genome coverage and peak identified

`ggcoverage` utilizes `ggplot2` plotting system, so its usage is ggplot2-style!

-------------

# RNA-seq data
## Load the data
The RNA-seq data used here are from [Transcription profiling by high throughput sequencing of HNRNPC knockdown and control HeLa cells](https://bioconductor.org/packages/release/data/experiment/html/RNAseqData.HNRNPC.bam.chr14.html), we select four sample to use as example: ERR127307_chr14, ERR127306_chr14, ERR127303_chr14, ERR127302_chr14, and all bam files are converted to bigwig file with [deeptools](https://deeptools.readthedocs.io/en/develop/).

Load metadata:
```{r load_metadata}
# load metadata
meta.file <- system.file("extdata", "RNA-seq", "meta_info.csv", package = "ggcoverage")
sample.meta = read.csv(meta.file)
sample.meta
```

Load track files:
```{r load_track}
# track folder
track.folder = system.file("extdata", "RNA-seq", package = "ggcoverage")
# load bigwig file
track.df = LoadTrackFile(track.folder = track.folder, format = "bw",
                         meta.info = sample.meta)
# check data
head(track.df)
```

Prepare mark region:
```{r prepare_mark}
# create mark region
mark.region=data.frame(start=c(21678900,21732001,21737590),
                       end=c(21679900,21732400,21737650),
                       label=c("M1", "M2", "M3"))
# check data
mark.region
```

Load GTF file:
```{r load_gtf}
gtf.file = system.file("extdata", "used_hg19.gtf", package = "ggcoverage")
gtf.gr = rtracklayer::import.gff(con = gtf.file, format = 'gtf')
```

-------------

## Basic coverage
```{r basic_coverage, warning=FALSE, fig.height = 6, fig.width = 12, fig.align = "center"}
basic.coverage = ggcoverage(data = track.df, color = "auto", 
                            mark.region = mark.region, range.position = "out")
basic.coverage
```

You can also change Y axis style:
```{r basic_coverage_2, warning=FALSE, fig.height = 6, fig.width = 12, fig.align = "center"}
basic.coverage = ggcoverage(data = track.df, color = "auto", 
                            mark.region = mark.region, range.position = "in")
basic.coverage
```

-------------

## Add gene annotation
```{r gene_coverage, warning=FALSE, fig.height = 8, fig.width = 12, fig.align = "center"}
basic.coverage + 
  geom_gene(gtf.gr=gtf.gr)
```

-------------

## Add transcript annotation
```{r transcript_coverage, warning=FALSE, fig.height = 12, fig.width = 12, fig.align = "center"}
basic.coverage + 
  geom_transcript(gtf.gr=gtf.gr,label.vjust = 1.5)
```

-------------

## Add ideogram
```{r ideogram_coverage_1, warning=FALSE, fig.height = 10, fig.width = 12, fig.align = "center"}
basic.coverage +
  geom_gene(gtf.gr=gtf.gr) +
  geom_ideogram(genome = "hg19",plot.space = 0)
```

```{r ideogram_coverage_2, warning=FALSE, fig.height = 14, fig.width = 12, fig.align = "center"}
basic.coverage +
  geom_transcript(gtf.gr=gtf.gr,label.vjust = 1.5) +
  geom_ideogram(genome = "hg19",plot.space = 0)
```

-------------

# DNA-seq data
## CNV
### Load the data
The DNA-seq data used here are from [Copy number work flow](http://bioconductor.org/help/course-materials/2014/SeattleOct2014/B02.2.3_CopyNumber.html), we select tumor sample, and get bin counts with `cn.mops::getReadCountsFromBAM` with `WL` 1000.

```{r load_bin_counts}
# track file
track.file = system.file("extdata", "DNA-seq", "CNV_example.txt", package = "ggcoverage")
track.df = read.table(track.file, header = TRUE)
# check data
head(track.df)
```

### Basic coverage
```{r basic_coverage_dna, warning=FALSE, fig.height = 6, fig.width = 12, fig.align = "center"}
basic.coverage = ggcoverage(data = track.df,color = NULL, mark.region = NULL,
                            region = 'chr4:61750000-62,700,000', range.position = "out")
basic.coverage
```

### Add annotations
Add **GC**, **ideogram** and **gene** annotations.

```{r gc_coverage, warning=FALSE, fig.height = 10, fig.width = 12, fig.align = "center"}
# load genome data
library("BSgenome.Hsapiens.UCSC.hg19")
# create plot
basic.coverage +
  geom_gc(bs.fa.seq=BSgenome.Hsapiens.UCSC.hg19) +
  geom_gene(gtf.gr=gtf.gr) +
  geom_ideogram(genome = "hg19")
```

---------------------

## Single-nucleotide level
### Load the data
```{r load_single_nuc}
# prepare sample metadata
sample.meta <- data.frame(
  SampleName = c("tumorA.chr4.selected"),
  Type = c("tumorA"),
  Group = c("tumorA")
)
# load bam file
bam.file = system.file("extdata", "DNA-seq", "tumorA.chr4.selected.bam", package = "ggcoverage")
track.df <- LoadTrackFile(
  track.file = bam.file,
  meta.info = sample.meta,
  single.nuc=TRUE, single.nuc.region="chr4:62474235-62474295"
)
head(track.df)
```

### Add base and amino acid annotation
```{r base_aa_coverage, warning=FALSE, fig.height = 10, fig.width = 12, fig.align = "center"}
ggcoverage(data = track.df, color = "grey", range.position = "out", single.nuc=T, rect.color = "white") +
  geom_base(bam.file = bam.file,
            bs.fa.seq = BSgenome.Hsapiens.UCSC.hg19) +
  geom_ideogram(genome = "hg19",plot.space = 0)
```

---------------------

# ChIP-seq data
The ChIP-seq data used here are from [DiffBind](https://bioconductor.org/packages/release/bioc/html/DiffBind.html), I select four sample to use as example: Chr18_MCF7_input, Chr18_MCF7_ER_1, Chr18_MCF7_ER_3, Chr18_MCF7_ER_2, and all bam files are converted to bigwig file with [deeptools](https://deeptools.readthedocs.io/en/develop/).

Create metadata:
```{r load_metadata_chip}
# load metadata
sample.meta = data.frame(SampleName=c('Chr18_MCF7_ER_1','Chr18_MCF7_ER_2','Chr18_MCF7_ER_3','Chr18_MCF7_input'),
                         Type = c("MCF7_ER_1","MCF7_ER_2","MCF7_ER_3","MCF7_input"),
                         Group = c("IP", "IP", "IP", "Input"))
sample.meta
```

Load track files:
```{r load_track_chip}
# track folder
track.folder = system.file("extdata", "ChIP-seq", package = "ggcoverage")
# load bigwig file
track.df = LoadTrackFile(track.folder = track.folder, format = "bw",
                         meta.info = sample.meta)
# check data
head(track.df)
```

Prepare mark region:
```{r prepare_mark_chip}
# create mark region
mark.region=data.frame(start=c(76822533),
                       end=c(76823743),
                       label=c("Promoter"))
# check data
mark.region
```

-------------

## Basic coverage
```{r basic_coverage_chip, warning=FALSE, fig.height = 6, fig.width = 12, fig.align = "center"}
basic.coverage = ggcoverage(data = track.df, color = "auto", region = "chr18:76822285-76900000", 
                            mark.region=mark.region, show.mark.label = FALSE)
basic.coverage
```

-------------

## Add annotations
Add **gene**, **ideogram** and **peak** annotations. To create peak annotation, we first **get consensus peaks** with [MSPC](https://github.com/Genometric/MSPC), you can also use DEbChIP's `GetConsensusPeak` (`MSPC`'s wrapper) to do this.

```{r peak_coverage, warning=FALSE, fig.height = 10, fig.width = 12, fig.align = "center"}
# get consensus peak file
peak.file = system.file("extdata", "ChIP-seq", "consensus.peak", package = "ggcoverage")

basic.coverage +
  geom_gene(gtf.gr=gtf.gr) +
  geom_peak(bed.file = peak.file) +
  geom_ideogram(genome = "hg19",plot.space = 0)
```

---------------------

# Session info
```{r session}
sessionInfo()
```










